Distribution-oblivious Programming with Dish

### 2019-11-22 pldi/draft.tex

\section{Intermediate Representation/Dataflow graph}
\label{ir}

\kk{Inline listing?}

\kk{Try to be more formal in the section, by using the notation that
  is defined in the nodes paragraph.}

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable
representation. As mentioned in \ref{}, the two main abstractions of
the shell, are files that contain data, and commands that communicate
through these files. We propose a slight variation of the standard
dataflow graph, that is commonly found in stream processing systems,
as the representation of a distributed shell computation. The main
benefit of the dataflow graph model is that it clearly exposes
operator level parallelism \kk{Is this a common term? Is there any
  definitive citation for it?}, as different nodes can be independent
processing workers. In our case each node represents a command, and
each edge represents a file.

\kk{Before that I have to mention that the basic data type $D$ is a
  line of characters. If we move the edges section before the nodes
  that will be fine.}

@@

Nodes $f$ of the graph represent functions from one (possibly empty)
input stream to an output stream $f : D^* \rightarrow D^*$.  This
representation captures the majority of shell commands. The nodes $f$
are monotonic, meaning that they cannot retract their output. More
formally, if $f(x) = y$, then for any $x'$, there exists $y'$ such
that $f(x.x') = y.y'$, where $.$ is standard concatenation and will be
omitted from now on. An example of a node is the following command
\kk{Does it need ref? Also maybe this example is too trivial and can
  be removed.}:

@@

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation though is that commands in the shell might read their
input stream from a sequence of different sources. An illustrative
example is the command \kk{Inline listing}cat which reads its input
from a combination of files and its standard input.

@@

\kk{Should I reference that this example was takes from the cat
  manpage?} The example above reads file \texttt{x} until it
encounters the end of file, then reads from \texttt{stdin}, and then
reads from \texttt{y}. In order to handle this particularity, nodes in
the dataflow graph can have many incoming edges, that are clearly
ordered in a sequence. Note that this comes in contrast with standard
dataflow graph representations where multiple incoming edges either
mean zipping the input to pairs, or arbitraliy interleaving the
different inputs.

\kk{A note about nodes and commands that can write and read from more
  files that the ones in their streams}

Note that nodes can read from more files than just the files of their
stream (or respectively write to more files that their output stream),
however these files are static (e.g. used for configuration or logging
\kk{Would an example here help}) and do not represent streams of data,
and therefore are not considered part of the dataflow graph. Because
of the assumption in section Front End \ref{}, that each file exists
only once in the dataflow graph, we can safely assume that reading and
writing to these static files does not interfere with the distributed
implementation, and doesn't alter the behaviour of the program.

@@

\kk{Maybe edges has to be moved before nodes, so that we can define
  the line data type}

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D*$, where $D$ represents a line, that is a sequence of characters
followed by \verb|\n| \kk{or EOF?}. They could either refer to a
specific named file in the file system, or just be FIFO pipes that
used for interprocess communication \kk{Are there any other file
  types, such as URLs, ...?}.

\kk{Show an example pipe and its graph}.

In order to expose operator parallelism with the output of a command,
files that don't point to a resource can have an upper bound on the
lines that they can tranfer. The following example illustrates why
this is essential.

\kk{An example of a command that has two outputs and one goes to one
  command, while another output goes to a different one. Maybe only
  show the graph here and not the command itself.}

On the other hand, some forms of data parallelism can be exposed when
knowing the size of the input files. As mentioned in \ref{} some pure
commands (such as cat -n) only need line information to become
stateless, and knowing the size of a file could allow the system to
split it in different chunks that can be processed independently. To
account for that, edges that refer to an input resource contain the
number of lines of the file that they refer to. \kk{At the moment this
  is not supported by our real system. The whole paragraph above can
  go if we decide that it is not a good idea to talk about it now.}

@@

\kk{(Maybe) Depending on which of the two paragraphs goes last
  (nodes/edges), I have to explain there that the last incoming edge
  to a node must be unbounded (if it doesn't point to a resource). I
  am actually not sure whether I should include this here.}

@@

@@

\kk{The following assumption should probably be mentioned in the
  categories, since a command that is non-deterministic should not be
  categorized to be neither stateless or pure. Assumption: How can we
  make sure that nodes in the IR are deterministic. Do we make some
  assumption that enforces this? I guess that commands that are nodes
  in the IR are all deterministic?  That is stateless and pure
  commands are deterministic.}

%% \kk{All of our assumptions and study are based on the POSIX
%%   standard. Reference?}

\section{Distribution Planner}

\kk{I think that there is not enough content for the planner section
  and it should probably just be called graph transformations. Then we
  could have a small subsection at the end about how the planner
  applies transformations iteratively starting from the source of the
  graph and pushing them until the end.}

As mentioned in \Cref{ir}, the dataflow graph exposes operator level
parallelism. Different nodes in the graph can be executed
concurrently, communicating through the edges. Operator level
parallelism is already exposed by the shell, by using pipes \texttt{|}
and the background operator \texttt{\&}. Unfortunately, simple
operator level parallelism leaves a lot to be desired, since scaling a
computation requires manually introducing more operators.

In order to adequately utilize the available computational resources,
a distributed implementation should exploit data parallelism. Data
parallelism is based on the observation that some operations produce
the same result when performed independently on subsets of the input
data, merging the results afterwards. Data level parallelism is the
basic optimization that is used by popular distributed processing
systems such as MapReduce\kk{cite} and Apache Spark\kk{cite}.

\subsection{Graph transformations}

\kk{This section could be moved to the intermediate representation,
  leaving a small section about the iterative process of the planner
  in the implementation/runtime section}

In our case, this is achieved by iteratively applying graph
transformations that preserve correctness, while exposing
parallelization opportunities. We illustrate that with a motivating
example; consider the following program and its dataflow graph.

\begin{lstlisting}[language=sh, float=h, numbers=none, escapeinside={($}{$)}]
 grep ``foo'' f1 f2 // Add its graph
\end{lstlisting}

\noindent
In this example, \texttt{grep} is applied to files f1 and f2, and its
result is sent to standard output. However, the result would be the
same if we concatenated the results of \texttt{grep ``foo'' f1} and
\texttt{grep ``foo'' f2}. This intuition can be generalized to all
stateless commands, as they are all essentially ``linear'' with
respect to concatenation and the following equation holds.

\[
\forall f, \mathrm{stateless}(f) \implies f(x.x') = f(x).f(x')
\]

\kk{Complete this story with pure, after we have written the section
  about command categories.}

\kk{Mention the assumptions that need to hold for the graph
  transformations to be valid. Commands must be deterministic, they
  must not do any other side effect (such as writing to other files,
  sending signals, etc). However, these assumptions must already be
  checked when the developer designates the categories.}

\subsection{Planner Algorithm}

The \texttt{Dish} planner extends the transformations that were
described above with edge splits in order to allow for the above
transformations to be performed. More precisely, it splits input files
and graph edges into several chunks so that they can be processed in
parallel by the next levels.

\kk{Explain the greedy algorithm that our planner performs. It splits
  input files, and then pushes parallelization through the graph,
  starting from the source nodes.}

@@

\subsection{Mapping operators to nodes}

\kk{I think we should actually remove this completely.}

Explain the simple algorithm that we used to minimize intra node
communication, that tries to map contiguous parts of the graph to the
same node. We could do this by having an algorithm that minimizes the
number of cuts or something.

We don't try to make a proper planner, as there is a lot of work on
operator placement etc. that we can borrow from.

@@

\kk{Rename this section to panner activators}

@@

\paragraph{Merge and Split}

\kk{Mergers and splitters, and how they are implemented using shell
  constructs (cat and head). This subsection might be a better fit in
  some other section, maybe something related to the planner or the
  implementation, as it is a very elegant way to implement splitters
  and mergers while staying completely inside the shell programming
  model.}

It is common in the dataflow graph for a node to have many incoming
edges, or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might read their input from one file
(e.g. stdin). In this case, the distributed implementation needs to
concatenate input files, and split output files. Here is an example:

@@

\kk{This transition needs to be stronger} Luckily, the shell offers
primitives for file manipulations that can be used to implement a
merger and a splitter. More concretely a merger can be implemented by
a simple \verb|cat|. Given a set of input files \verb|in1, in2, ...|
and an output file \verb|out|, a merger can be implemented as:

@@

\noindent
On the other hand, given an input file \verb|in| that is a FIFO pipe,
and two output files \verb|out1, out2| the first of which is bounded
to \verb|N| lines, a 2-splitter can be implemented as:

@@

\kk{Do I also need to include a splitter for normal files?}

>>>>

\section{Dataflow Graph Model}
\label{ir}

\TODO{Replace texttt with inline listings?}

In order to effectively apply optimizations and distribute a shell
script, it first has to be translated to a manipulable
representation. As mentioned in \ref{}, the two main abstractions of
the shell, are files that contain data, and commands that communicate
through these files. We propose a slight variation of the standard
dataflow graph model, that is commonly found in stream processing
systems, as the representation of a distributed shell computation. The
main benefit of the dataflow graph model is that it clearly exposes
task parallelism, as different nodes can be independent processing
workers. In our case each node represents a command, and each edge
represents a file.

\subsection{Graph Components}

@@

Nodes $f$ of the graph represent functions from one (possibly empty)
input stream to an output stream $f : D^* \rightarrow D^*$. $D$
represents the basic data type of a line of characters. This
representation captures the majority of shell commands. The nodes are
monotonic, meaning that they cannot retract their output. More
formally, $\forall f, x, y, f(x) = y$, then for any $x'$, there exists
$y'$ such that $f(x.x') = y.y'$, where $.$ is standard concatenation
and will be omitted when obvious. An example of a node is the
following command \kk{Does it need ref? Also maybe this example is too
  trivial and can be removed.}:

@@

\noindent
This command takes a stream of lines from its standard input and
returns a stream of lines on its standard output. An important
observation is that commands in the shell might read their input
stream from a sequence of different sources. An illustrative example
is the command \texttt{cat} which reads its input from a combination
of files and its standard input.

@@

The example above reads file \texttt{x} until it encounters the EOF
character, then reads from \texttt{stdin}, and then reads from
\texttt{y}. In order to handle this particularity, nodes in the
dataflow graph can have many incoming edges, that are ordered in a
sequence. Note that this does not correspond to standard dataflow
graph representations where the semantics of multiple incoming edges
is either zipping the inputs in pairs, or arbitraliy interleaving them.

\TODO{Is the following paragraph best suited here? It describes an
  assumption that we make about commands, to guarantee the validity of
  our transformations. The assumption is needed due to the limitation
  of the programming model of one input one output per command.}

Note that nodes can read from more files than just the files of their
stream (or respectively write to more files that their output stream),
however these files are static (e.g. used for configuration or logging
\kk{Would an example here help}) and do not represent streams of data,
and therefore are not considered part of the dataflow graph. We assume
that these static files can only be accessed by one node, thus
ensuring that they do not interfere with the distributed
implementation.
%% Because of the assumption in section Front End \ref{}, that each file
%% exists only once in the dataflow graph, we can safely assume that
%% reading and writing to these static files does not interfere with the
%% distributed implementation, and doesn't alter the behaviour of the
%% program.

@@

Edges in the dataflow graph represent files, the basic data
abstraction of the shell. They are used as communication channels
between nodes in the graph, and as the input or output of the whole
graph. Edges are represented as possibly unbounded streams of type
$D*$. They could either refer to a specific named file in the file
system, or just be FIFO pipes that used for interprocess communication
\TODO{Are there any other file types, such as URLs, ...?}.

\TODO{Show an example pipe and its graph}.

In order to expose data parallelism from the output of a command,
files that don't point to a resource can have an upper bound on the
lines that they can tranfer. The following example illustrates why
this is essential.

\TODO{Add an example of a command that has two outputs and one goes to
  one command, while another output goes to a different one. Maybe
  only show the graph here and not the command itself.}

On the other hand, some forms of data parallelism can be exposed when
knowing the size of the input files. As mentioned in \ref{} some pure
commands (such as cat -n) only need line information to become
stateless, and knowing the size of a file could allow the system to
split it in different chunks that can be processed independently. To
account for that, edges that refer to an input resource contain the
number of lines of the file that they refer to.

@@

@@

\subsection{Graph transformations}
\label{ir:transformations}

As mentioned above, the dataflow graph exposes task
parallelism. Different nodes in the graph can be executed
concurrently, communicating through the edges. Task level parallelism
is already exposed by the shell, by using pipes \texttt{|} and the
background operator \texttt{\&}. Unfortunately, simple task
parallelism leaves a lot to be desired, since scaling a computation
requires manually introducing more operators.

In order to adequately utilize the available computational resources,
a distributed implementation should exploit data parallelism. Data
parallelism is based on the observation that some operations produce
the same result when performed independently on subsets of the input
data, merging the results afterwards. This is the basic optimization
that is used by popular distributed processing systems such as
MapReduce\kk{cite} and Apache Spark\kk{cite}. \todo{Unfortunately, the
  dataflow graph model of most of these systems is not expressive
  enough to represent data parallelism, so the distributed generations
  that they generate, can not be represented in their dataflow model.}
\kk{This last sentence is extremely important and we should make it
  shine.}

In contrast, our dataflow model inherently supports the data
parallelism that is found in shell programs. More precisely, data
parallelism can be easily exposed as task parallelism in our model. A
very important consequence of this is that we can iteratively apply
optimizations that expose data parallelism, or even revert them when
the resulting dataflow graph is too wide. This flexibility enables
unified reasoning and implementation of the unoptimized and optimized
version of the dataflow graph. \kk{This is exciting, and maybe it
  should be moved up (or rephrased in the introduction)}

Before describing the different types of parallelization
optimizations, we formalize our intuition about stateless and pure
commands. Recall the example \ref{} from section Categories
\ref{}. This example shows the stateless command x being applied to
...\kk{Complete this after we have an example}. Formally, a commmand
$f$ is stateless if it is ``linear'' with respect to concatenation,
i.e. the following equation holds:

\[
\forall x, x', f(x.x') = f(x).f(x')
\]

Similarly, every pure command $f$ that can be written as a map $m$ and a
reduce $r$ step satisfies the following:

\[
\forall x, x', f(x.x') = r(m(x).m(x'))
\]

Based on these equations, we can define a maximal node split operation
on a stateless node $v$ with $n$ input ordered edges and an output
edge to node $v'$, to replace $v$ with $n$ new nodes, giving each one
of them one of the incoming edges of $v$, and routing them in order to
point to $v'$. The maximal node split operation can similarly extended
for pure nodes.

\kk{If there is time I can work out a formal definition and a proof
  sketch why this transformation preserves the output of the dataflow
  graph}.

%% \begin{definition}
%% Given a dataflow graph $G = (V, E, O)$, where $V$ is a set of nodes
%% representing commands, $E$ is a set of edges representing files, and
%% $O$ is a function from $V \cup v_{out}$ to a total order of incoming
%% edges. $v_{out}$ represents a concatenation of all the outputs of the
%% dataflow graph. We represent the total order for a node $v$ as
%% $<_v$. Given a node a node $v$
%% with input edges $ie = \{ i_1 = (v_{i_1}, v, 1), i_2, ..., i_n =
%% (v_{i_n}, v, n) \}$ and output edge $o = (v, v_o, a)$, we define a
%% complete node split $s(v, G) = (V', E')$ where $V' = V - v \cup \{
%% v_1, ..., v_n \}$ and $E' = E - ie \cup \{(v_{i_1}, v_1, 1), ...,
%% (v_{i_n}, v_n, 1) \} - o \cup \{ (v_1, v_o\}$
%% \end{definition}

%% \begin{lemma}
%% tofill
%% \end{lemma}

In addition to node splits, we can also perform edge splits in order
to increase the number of incoming edges for some nodes to expose
possible data parallelism. If the edge refers to a real file, then the
edge split represents splitting the real file in chunks, while in
cases where the edge connects two nodes, it is split to several edges
with a bounded number of lines. Note that the final output edge of a
valid node in the graph must always be unbounded if it does not refer
to a file.

\kk{Remember to mention the assumptions that need to hold for the
  graph transformations to be valid in the Command categories
  section. Commands must be deterministic, they must not do any other
  side effect (such as writing to other files, sending signals,
  etc). However, these assumptions must already be checked when the
  developer designates the categories.}

@@

\section{Dish Implementation}

\kk{I think that this section should be merged with the front end}

\paragraph{Dish Optimizer}

The \texttt{Dish} optimizer is given a dataflow graph representing a
shell program and outputs an optimized dataflow graph, where data
parallelism opportunities have been exploited. More concretely, the
optimizer starts by splitting input files in several chunks, given the
desirable output dataflow graph width. It then starts from the source
nodes of the graph, iteratively performing the graph transformations
that were described in \Cref{ir:transformations}. If the incoming
edges of the node are less than the desirable graph width, it performs
edge splits, and then it performs maximal node splits. The final graph
can then be implemented by spawning a process for each node, and
redirecting the outputs according to the graph edges.

@@

%% \subsection{Mapping operators to nodes}

%% \kk{I don't think we should mention this.}

%% Explain the simple algorithm that we used to minimize intra node
%% communication, that tries to map contiguous parts of the graph to the
%% same node. We could do this by having an algorithm that minimizes the
%% number of cuts or something.

%% We don't try to make a proper planner, as there is a lot of work on
%% operator placement etc. that we can borrow from.

@@

\TODO{Rename this section to planner activators and write it after I
  discuss with Nikos.}

@@

\subsection{Merge and Split Implementation}

\TODO{Niko, could you make this a bit stronger?}

It is common in the dataflow graph for a node to have many incoming
edges, or many outgoing edges (when all but the last outgoing edges
are bounded). However some nodes, might only read their input from one
source (e.g. stdin). In this case, the distributed implementation
needs to concatenate input files (or conversely split output
files). Here is an example:

@@

We address this issue using primitive shell constructs for file
manipulation. More concretely a merger can be implemented by a simple
\verb|cat|. Given a set of input files \verb|in1, in2, ...| and an
output file \verb|out|, a merger can be implemented as:

@@

\noindent
\TODO{Change that with the new splitter}
On the other hand, given an input file \verb|in| that is a FIFO pipe,
and two output files \verb|out1, out2| the first of which is bounded
to \verb|N| lines, a 2-splitter can be implemented as:

@@
